---
title: "9. Intro Statistics"
author: "David LeBauer"
date: '2017-06-05'
slug: intro-statistics
categories: []
tags: [statistics,r]
---


## Summary statistics

Objectives:
* learn to summarize data with basic measures of central tendancy and spread
* learn to fit parametric distributions to data

### Data analysis: Sorghum height data

```{r}
knitr::opts_chunk$set(cache = TRUE)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())

#bety_src <- RPostgreSQL::dbConnect(odbc::odbc(), "TERRA-REF traits copy BETYdb")
#bety_src <- src_postgres(dbname = "bety", password = 'bety', host = 'bety.terraref', user = 'bety', port = 5432)
bety_src <- src_postgres(dbname = "bety", 
                         password = 'DelchevskoOro', 
                         host = 'bety6.ncsa.illinois.edu', 
                         user = 'viewer', 
                         port = 5432)
#bety_src <- RPostgreSQL::dbConnect(odbc::odbc(), "TERRA-REF traits viewer")

```

```{r complex-query}
treatments <- tbl(bety_src, 'treatments') %>% 
  dplyr::select(treatment_id = id , name, definition, control)

managements_treatments <- tbl(bety_src, 'managements_treatments')

managements <- tbl(bety_src, 'managements') %>% 
  filter(mgmttype %in% c('Fertilization_N', 'Planting', 'Irrigation')) %>% 
  dplyr::select(management_id = id, date, mgmttype, level, units) %>% 
  left_join(managements_treatments, by = 'management_id') %>% 
  left_join(treatments, by = 'treatment_id')

planting <-managements %>% 
  filter(mgmttype == "Planting") %>% 
  dplyr::select(treatment_id, planting_date = date, nrate = level)

canopy_height <- tbl(bety_src, 'traits_and_yields_view') %>% 
  filter(trait == 'canopy_height') %>% 
  left_join(planting, by = 'treatment_id') %>% 
  collect
```


```{r}
ggplot(data = canopy_height) + 
  geom_histogram(aes(x = mean), binwidth = 10) 


```


## Basic Statistics

These are some key statistics that are useful for describing a random variable $X$:

* Mean $\bar{x}=\frac{1}{n}\sum_{i=1}^{n}{x_i}$
* [Median](https://en.wikipedia.org/wiki/Median) 
* Variance $\rm{Var}(X) = \frac{1}{n}\sum_{i=1}^{n}{(x_i-\mu)^2}$
* Standard Deviation $\rm{SD}(X) = \sqrt{\rm{Var}(X)}$
* Coefficient of Variance $\rm{CV}(X) = \frac{\rm{SD}(X)}{\bar{x}}$
* [Skewness](https://en.wikipedia.org/wiki/Skewness): does the distribution skew left, or right? 
   * when is median > mean?
* [Kurtosis](https://en.wikipedia.org/wiki/Kurtosis): how fat are the tails of the distribution? 

![https://en.wikipedia.org/wiki/Skewness#/media/File:Negative_and_positive_skew_diagrams_(English).svg] [Rodolfo Hermans](https://en.wikipedia.org/wiki/User:Rodolfo_Hermans)



```{r}
x <- canopy_height$mean
mean(x)
var(x)
sd(x)

cv <- function(x){
  cv <- sd(x)/mean(x)
  return(cv)
}

cv(x)

### need to install moments package to compute skewness and kurtosis
# install.packages('moments')
moments::skewness(x)
moments::kurtosis(x)
```

```{r}
height_season1 <- canopy_height %>% filter(grepl('Season 1', sitename))

ggplot(data = height_season1, aes(date, mean)) +
  geom_point(alpha = 0.25)


```

### Fitting distributions to data

Probability distributions provide a convenient way of describing a random variable. In many cases, the parameters can be interpereted in a meaningful way (e.g. shape, rate, scale).


```{r}
#install.packages('fitdistrplus')

library(fitdistrplus)

descdist(x)

plotdist(x)

w <- fitdist(x, 'weibull')
ln <- fitdist(x, 'lnorm')
g <- fitdist(x, 'gamma')
n <- fitdist(x, 'norm')

which.min(c(w$aic, ln$aic, g$aic, n$aic))

plot(g)
plot(ln)
plot(n)
plot(w)

hist(x, probability = TRUE, ylim = c(0, 0.012))
lines(sort(x), dgamma(sort(x), g$estimate['shape'], g$estimate['rate']))
lines(sort(x), dlnorm(sort(x), ln$estimate['meanlog'], ln$estimate['sdlog']), col = 2)
lines(sort(x), dnorm(sort(x), n$estimate['mean'], n$estimate['sd']), col = 3)
lines(sort(x), dnorm(sort(x), w$estimate['shape'], n$estimate['scale']), col = 3)

```

```{r}
mean(x)
plot(x)
lm(x~1)
```


## Regression 

Objectives:

* learn to convert deterministic functions into statistical models
* fit parameters
* evaluate key assumptions
* learn to interpret model summaries / parameters / output
* use model comparison to test hypotheses 


We are starting with hypotheses / preidctions that the yield of two perennial grasses, _Miscanthus_ spp. and Switchgrass (_Panicum spp._)) depend on age, fertilization rate, and genotype. We expect that the functional form of the response to nitrogen and age will either be monotonically increasing, asymptotic, or hump-shaped. This follows [LeBauer et al 2017](http://onlinelibrary.wiley.com/doi/10.1111/gcbb.12420/full) using a subset of the data that is balanced data from Miscanthus and Switchgrass field trials within Il and across the US ([Arundale et al 2012](http://hdl.handle.net/2142/34422) and [Arundale et al 2014](http://onlinelibrary.wiley.com/doi/10.1111/gcbb.12077/full))




```{r}
library(traits)

options(
  betydb_url = "https://betydb.org",
  betydb_api_version = 'beta')

yields <- betydb_query(table = 'search', 
                       trait = 'Ayield', 
                       limit = "none")

grass_yields <- yields %>% 
  filter(genus %in% c('Miscanthus', 'Panicum')) %>% 
  dplyr::rename(yield_annual = mean)


```

We are going to use a version of this dataset that already has climate data: MAP (Mean Annual Precipitation) and Mean Annual Temperature (MAT)

```{r all.yieldata, fig.width=9, fig.height=9}

grass_yields <- read.csv('data/grass_yield.csv')

#lattice::splom(grass_yields)
#pairs(grass_yields, pch='.')

#grass_yields <- grass_yields %>% filter(genus %in% c("Miscanthus", "Panicum"))
```


```{r}
h0 <- lm(yield_annual ~ 1, data = grass_yields)

plot(h0$fitted.values, h0$residuals)
hist(h0$residuals)


plot(h0, which = c(1,2))
```

### Data Transformation 

```{r}
grass_yields <- grass_yields %>% mutate(log_yield = log10(yield_annual + 1))
```

Why add 1 to `yield_annual`? How does this affect interpretation?

```{r}
h0 <- lm(log_yield ~ 1, data = as.data.frame(grass_yields))

plot(h0$fitted.values, h0$residuals)
hist(h0$residuals)
```

### extending the models
```{r}
mod1 <- lm(yield_annual ~ genus, data = grass_yields)
summary(mod1)

methods(class = class(mod1))
```


### Assumptions

```{r}
par(mar = c(4, 4, 2, 2), mfrow = c(1, 2)) #optional
plot(mod1, which = c(1, 2)) # "which" argument optional
```

1. residuals vs fitted: are residuals distributed consistently along axis of predictor variables?
2. qqplot: are errors distributed normally?

see http://data.library.virginia.edu/diagnostic-plots/

### Extending the model

```{r}

mod3 <- lm(yield_annual ~ genus + mat + map, data = grass_yields)
mod6 <- lm(yield_annual ~ genus + mat + map + age + fertilizer_n + planting_density, data = grass_yields)

summary(mod3)

```

### Comparing models

```{r}
mod4 <- lm(yield_annual ~ genus + mat + map + age, data = grass_yields)
AIC(mod1, mod4, mod6)

BIC(mod1, mod4)
```

### Interactions

```{r}
mod4 <- lm(yield_annual ~ genus * age + mat + map, data = grass_yields)
```

how do you know which interactions to include?


```{r}
anova(mod1, mod4)

```
refining the model

```{r}
library(MASS)
mod4 <- lm(yield_annual ~ genus * age + genus * mat + genus * map, data = grass_yields)

```

```{r eval=FALSE}
ggplot(data = grass_yields) +
  geom_histogram(aes(mean, y = ..density.., fill = genus), position = 'dodge', binwidth = 5)
```


```{r eval = FALSE}
x <- grass_yields$yield_annual
mean(x)
var(x)
sd(x)
cv <- function(x){
  sd(x)/mean(x)
}
cv(x)

library(ggplot2)
ggplot() +
  geom_histogram(aes(x), bins = 25)

u <- runif(1000)
sum(u < 0.11)
z <- ifelse(u < 0.11, runif(1000), rgamma(1000, 6, 1))

yield <- grass_yields$yield_annual
x <- seq(0, 50, by = 0.1)

hist(yield, breaks = 25, probability = TRUE)
lines(x, dnorm(x, mean = mean(yield), sd = sd(yield)))

yield_subset <- yield[yield>1]
hist(yield_subset, breaks = 25, probability = TRUE)
lines(x, dnorm(x, mean = mean(yield), sd = sd(yield)))

w <- fitdist(yield_subset, distr = 'weibull')
ln <- fitdist(yield_subset, distr = 'lnorm')
g <- fitdist(yield_subset, distr = 'gamma')
n <- fitdist(yield_subset, distr = 'norm')

plot(g)
plot(n)
plot(ln)

names(g)
g$estimate

n$loglik
g$loglik

x <- yield_subset
hist(x, probability = TRUE, ylim = c(0, 0.10))
lines(sort(x), dgamma(sort(x), g$estimate['shape'], g$estimate['rate']))
lines(sort(x), dlnorm(sort(x), ln$estimate['meanlog'], ln$estimate['sdlog']), col = 2)
lines(sort(x), dnorm(sort(x), n$estimate['mean'], n$estimate['sd']), col = 3)


unique(grass_yields$genus)

grass_yields$genus <- ifelse(grass_yields$genus == 'Freedom-', 
                             "Miscanthus",
                             grass_yields$genus)
grass_yields <- grass_yields %>% filter(!genus == 'root-mis')

theme_set(theme_bw())
ggplot(data = grass_yields) +
  geom_histogram(aes(x = yield_annual), binwidth = 4) +
  facet_wrap(~genus)

ggplot(data = grass_yields) +
  geom_point(aes(x = year, y = yield_annual)) +
  facet_wrap(~genus)

ggplot(data = grass_yields) +
  geom_point(aes(x = age, y = yield_annual)) +
  facet_wrap(~genus)

ggplot(data = grass_yields) +
  geom_point(aes(x = year, y = map)) +
  facet_wrap(~lat)

ggplot(data = grass_yields) +
  geom_point(aes(x = year, y = mat)) +
  facet_wrap(~lat)

ggplot(data = grass_yields) +
  geom_point(aes(x = lon, y = lat, color = genus), alpha = 0.25, position = 'jitter')

lattice::splom(grass_yields)
lines(sort(x), dweibull(sort(x), w$estimate['shape'], w$estimate['scale']), col = 4)



h0 <- lm(yield_annual ~ 1, data = grass_yields)

hist(grass_yields$yield_annual)
abline(v = mean(grass_yields$yield_annual))

plot(h0, which = c(1))
plot(h0, which = c(2))

## log
## power (incl. sqrt)
## Box-Cox transform
## rank
library(dplyr)
grass_yields <- grass_yields %>% 
  mutate(sqrt_yield = sqrt(yield_annual))

h0 <- lm(sqrt_yield ~ 1, data = grass_yields)

plot(h0, which = c(1))
plot(h0, which = c(2))

h1 <- lm(sqrt_yield ~ 1 + genus, data = grass_yields)

plot(h1, which = c(1))
plot(h1, which = c(2))

summary(h1)

class(h1)
methods(class = 'lm')

plot(h1$fitted.values, h1$residuals)

tmp_grass <- cbind(grass_yields, residuals = h1$residuals)
h1.5 <- lm(residuals ~ fertilizer_n, data = tmp_grass)

summary(h1.5)

h2 <- lm(sqrt_yield ~ genus + fertilizer_n, 
         data = grass_yields)

summary(h2)

h3 <- lm(sqrt_yield ~ 1 + genus + fertilizer_n + genus : fertilizer_n, data = grass_yields)
# h3 <- lm(sqrt_yield ~ 1 + genus + fertilizer_n + genus : fertilizer_n, data = grass_yields)
# h3 <- lm(sqrt_yield ~ 1 + genus * fertilizer_n , data = grass_yields)
library(ggplot2)
ggplot(data = grass_yields) +
  geom_point(aes(fertilizer_n, sqrt_yield, color = genus )) +
  geom_abline(aes(slope = 0.0125, intercept = 3.2),color = 'pink') +
  geom_abline(aes(slope = 0.0125 - 0.0066, intercept = 3.2 - 0.65),color = 'blue')

ggplot(data = grass_yields) +
  geom_point(aes(fertilizer_n, sqrt_yield, color = genus )) +
  geom_abline(aes(slope = 0.0125, intercept = 3.2),color = 'pink') +
  geom_abline(aes(slope = 0.0125 - 0.0066, intercept = 3.2 - 0.65),color = 'blue')
summary(h3)

ggplot(data = grass_yields) +
  geom_point(aes(age, sqrt_yield, color = genus ))

h5 <- lm(sqrt_yield ~ 1+ genus * mat, 
         data = grass_yields )
summary(h5)

h6 <- lm(sqrt_yield ~ 1+ genus * mat + genus * fertilizer_n + genus *age, 
         data = grass_yields )

h7 <- lm(sqrt_yield ~ 1+ genus + genus , 
         data = grass_yields )
summary(h6)

plot(h6, which = c(1))

AIC(h0)

library(MASS)

stepAIC(h6, direction = 'both')

h8 <- lm(formula = sqrt_yield ~ genus + fertilizer_n + age + I(age^2) + genus:fertilizer_n + 
     genus:age, data = grass_yields)

plot(h8)

?stepAIC

lapply(list(h0, h1, h3, h5, h6, h8), AIC)

```


## Issues with stepAIC

https://stats.stackexchange.com/questions/20836/algorithms-for-automatic-model-selection


```{r}
grass_yields <- read.csv('data/grass_yield.csv') %>% 
  filter(genus %in% c("Miscanthus", "Panicum")) %>%
  mutate(sqrt_yield = sqrt(yield_annual), genus = as.character(genus))

```

### Model fitting

```{r caret, eval=FALSE}
install.packages('caret')
require(caret)

ctrl <- trainControl(method = "repeatedcv", repeats = 10, savePred = TRUE)

train_lm_yield <- train(yield_annual ~  genus * fertilizer_n + genus * poly(age, 2) + genus * mat + genus * map, 
                  data = grass_yields,
                  preProcess = 'BoxCox',
                  method = "lmStepAIC", 
                  direction = 'both',
                  trControl = ctrl)

class(train_lm_yield$finalModel$call)
```

Alternative: lasso
```{r eval = FALSE}

yield_lasso <- train(yield_annual ~  genus * fertilizer_n + genus * poly(age, 2) + genus * mat + genus * map, 
                  data = grass_yields,
                  preProcess = 'BoxCox',
                  method = "glmnet")

varImp(lasso_yield)

library(sjPlot)

sjPlot::sjp.lm(train_lm_yield$finalModel)
summary(train_lm_yield)

```

## Random Effects

```{r eval=FALSE}
## create a category for site
grass_yields <- grass_yields %>% 
  mutate(site = as.factor(lat))

library(lme4)

grass_re <- lme4::lmer(sqrt_yield ~ fertilizer_n + age + I(age^2) +
                       genus:fertilizer_n + genus:age + (1 | site), 
                  data = grass_yields)

sjPlot::sjp.lmer(grass_re)
```

### the Predict method

take an argument 'new data' and predict output from the model

```{r}

grass_lm <- lm(sqrt_yield ~ genus * fertilizer_n + poly(age, 2), data = grass_yields)

newdata <- expand.grid(fertilizer_n = 1:5*100, age = 5, genus = 'Miscanthus', year = 2000)

y_n <- predict(grass_lm, newdata = newdata)

plot(1:5*100, y_n)
```

The danger of 'out of sample prediction'

```{r}
newdata2 <- expand.grid(fertilizer_n = 100, age = 1:20, genus = 'Miscanthus', year = 2000)

y_age <- predict(grass_lm, newdata = newdata2)
plot(1:20, y_age)
```


#### References

Arundale R (2012) The higher productivity of the bioenergy feedstock Miscanthus x giganteus relative to Panicum virgatum is seen both into the long term and beyond Illinois (Doctoral dissertation, University of Illinois at Urbana-Champaign). http://hdl.handle.net/2142/34422.

Arundale RA, Dohleman FG, Heaton EA, Mcgrath JM, Voigt TB, Long SP (2014) Yields of Miscanthus × giganteus and Panicum virgatum decline with stand age in the Midwestern USA. Global Change Biology Bioenergy, 6, 1–13.
http://onlinelibrary.wiley.com/doi/10.1111/gcbb.12077/full

LeBauer, D., Kooper, R., Mulrooney, P., Rohde, S., Wang, D., Long, S. P. and Dietze, M. C. (2017), BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. GCB Bioenergy. doi:10.1111/gcbb.12420

### Fitting mathematical models


functional forms https://ms.mcmaster.ca/~bolker/emdbook/book.pdf


The following is an illustrative collection of the mathematical functions based on those described in Bolker 2009.

```{r eval = FALSE}
foo.null  <- lm (y ~ 1, ...)
foo.poly1 <- lm(y ~ poly(x,1),...)
foo.poly2 <- lm(y ~ poly(x,2)+poly(x2,2),...)
foo.poly3 <- lm(y ~ poly(x,3),...)
foo.mm <- nls ( y~ (a*x)/(b + x) , start = list(a=1, b=1),...)
foo.ne <- nls ( y ~ a*exp(-b*x) , start = list(a=1, b=1),...)
foo.ricker <- nls ( y ~ a*x*exp(-b*x) , start = list(a=1, b=1),...)
foo.powricker <- nls ( y ~ b * (x/a * exp(1 - x/a))^alpha , start = list(a=1, b=1, alpha = 1),...)
foo.modlog <- nls ( y ~ exp(eps * (phi - x))/(1+exp(beta * eps * ( phi - x)))
 , start = list(eps = 1, beta = 1, phi = 15),...)
foo.aov <- aov( y ~ x, ...)

```

### Now ... this leads us to mechanistic modeling


Next lecture on Crop modeling 

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vRkb_-BtuFGBh0erv_yuE6BApuZ0rVMti9NN2XmGWcT9iC-muHZo2MyCMno4HLmIcSGlkCfpSat2fVw/embed?start=false&loop=false&delayms=3000" frameborder="0" width="1280" height="749" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true"></iframe>
