---
title: Exploratory Data Analysis
author: "David LeBauer"
date: '2017-05-31'
slug: exploratory-data-analysis
categories: []
tags: []
---

> **Note**: Some of the commands below, particularly those accessing the terraref database from within RStudio, require being logged into the [NDS Analytics Workbench](http://www.terraref.ndslabs.org/#/). This requires signing up for the [TERRA REF Alpha User program](https://docs.google.com/forms/d/e/1FAIpQLScBsD042RrRok70BCGCRwARTcm9etvVHqvQaz1c5X7c5y0H3w/viewform?c=0&w=1).

## Objectives

* Learning how to access data on terraref database via PostgreSQL

* Learning how to manipulate dat using the `dplyr` library, in particular:
   * [filtering, subsetting, summarizing, new variables with dplyr](http://swcarpentry.github.io/r-novice-gapminder/13-dplyr/)
   * [converting data from wide to long with tidyr](http://swcarpentry.github.io/r-novice-gapminder/14-tidyr/)

## Accessing TERRA REF via PostgreSQL

We want to access the TERRA REF database. We will do it using the RPostgreSQL package:

```{r}
install.packages("RPostgreSQL")
```

You can then connect to the database using (**note:** this is one of those commands that require you to be working in the NDS Analytics Workbench):

```{r}
bety_src <- src_postgres(dbname = 'bety',
                password = 'bety',
                host = 'terra-bety.default',
                user = 'bety',
                port = 5432)
```

## The dplyr library

We will be using the library [dplyr](https://cran.rstudio.com/web/packages/dplyr/README.html), which provides helps with data manipulation.

```{r}
library(dplyr)
```

#### filtering / subsetting


Let's get some data:

```{r}
traits <- tbl(bety_src, 'traits', n=Inf)

variables <- tbl(bety_src, 'variables') %>%
  mutate(variable_id = id)

sites <- tbl(bety_src, 'sites') %>%
  mutate(site_id = id)
```

In the above code the symbol `%>%` is a pipe from the `dplyr` library, it passes the left hand side to the first argument slot of the right hand side.

Using `filter` we can pick what we see based on their values:

```{r}
traits_variables <- traits %>%
  left_join(variables, by = 'variable_id') %>%
  left_join(sites, by = 'site_id') %>%
  select(sitename, name, mean) %>%
  filter(name == 'canopy_height') %>%
  collect()

traits_variables_season_2 <- traits_variables %>%
      filter(grepl("Season 2",sitename))
```

### grouping and summarizing

One can combine the `group_by` and `summarize` functions to get a summary of the data we pulled:

```{r}
traits_variables_season_2 %>%
  group_by(date) %>%
  summarize(mean = mean(mean), n =n())

canopy_heights <- traits_variables_season_2 %>%
  group_by(date, cultivar_id) %>%
  summarize(mean = mean(mean), n =n())
```


## Plotting with ggplot

The `ggplot2` library (for more see [here](https://cran.r-project.org/web/packages/ggplot2/index.html)) can be used to create graphics.

```{r}
library(ggplot2)
```

An example using the data we pulled before:

```{r}
ggplot(data = traits_variables_season_2,
       aes(x = date, y = mean, group = cultivar_id)) +
  geom_smooth(se = FALSE, alpha = 0.25, size = 0.5) +
  geom_point(alpha = 0.25, size = 0.5, position = 'jitter')
```

Another example:

```{r}
### filter a subset of 25 genotypes
height_season_2_subset <- traits_variables_season_2 %>%
  filter(cultivar_id < min(traits_variables_season_2$cultivar_id)+25)

# how many unique values of cultivar_id?

#height_season_2_subset %>% group_by(cultivar_id, date) %>% summarize(n())


ggplot(data = height_season_2_subset,
       aes(x = date, y = mean, color = as.factor(cultivar_id))) +
  geom_smooth(se = FALSE, alpha = 0.25, size = 0.5) +
  geom_point(alpha = 0.25, size = 0.5)
```

### Faceting

Using `ggplot2` we can partition a plot into a matrix of panels. Here's an example:

```{r}
ggplot(data = height_season_2_subset,
       aes(x = date, y = mean)) +
  geom_smooth(se = FALSE) +
  geom_point() +
  facet_wrap(~cultivar_id) +
  ggthemes::theme_tufte()
```

### Other plots

Here is an example of a histogram:

```{r}
ggplot(data = height_season_2_subset,
       aes(mean, fill = date)) +
  geom_histogram() +
  facet_wrap(~cultivar_id) +
  theme_bw()
```

> Try making plots using the following functions:
> * `geom_boxplot`
> * `geom_violin`




### Visualization in R using ggplot

Some references for using ggplot:

* R for data science  [data-visualization](http://r4ds.had.co.nz/data-visualisation.html)
* [Software Carpentry 8](http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/)
