---
title: Stochastic Simulation
author: "David LeBauer"
date: '2017-06-06'
slug: stochastic-simulation
categories: []
tags: []
---

> **Note**: Some of the commands below, particularly those accessing the terraref database from within RStudio, require being logged into the [NDS Analytics Workbench](http://www.terraref.ndslabs.org/#/). This requires signing up for the [TERRA REF Alpha User program](https://docs.google.com/forms/d/e/1FAIpQLScBsD042RrRok70BCGCRwARTcm9etvVHqvQaz1c5X7c5y0H3w/viewform?c=0&w=1)

```{r}
library(ggplot2)
```


### Stochastic Simulation (Bolker Ch5)


A deterministic linear function:

$$y_\rm{det}=a+bx$$

Add normally distributed errors, modeling $\epsilon\sim(N(0,\sigma))$

$$Y\sim N(a+bx, \sigma)$$
Or equivalently:

$$Y\sim N(\mu, \sigma)$$

$$\mu = a+bx$$
Lets first plot the deterministic function as a line

```{r}
set.seed(10)
x <- 1:20
a <- 2
b <- 1
sd <- 2

y_det <- a + b * x

Y <- rnorm(length(x), mean = y_det, sd = sd)


plot(x, y_det, type = 'l')
points(x, Y)
lines(x, y_det+sd, lty = 2)
lines(x, y_det-sd, lty = 2)

```

```{r lm-fit-y-x}
fit <- lm(Y~x)
y_hat <- predict(fit)
ggplot() +
  geom_line(aes(x, y_det)) +
  geom_point(aes(x, Y)) +
  geom_line(aes(x, y_hat), linetype = 2)


a_hat <- vector()
b_hat <- vector()
for(i in 1:100){
  Y <- rnorm(length(x), mean = y_det, sd = 2)
  a[i] <- lm(Y~x)$coefficients["(Intercept)"]
  b[i] <- lm(Y~x)$coefficients["x"]
}

hist(a)
mean(a)
sd(a)

hist(b)
mean(b)
sd(b)
```


> What if $a$ and $b$ have the following disrtibutions?

$$a\sim N(2,0.1) \\\textrm{and}\\ b\sim N(1,0.01)\\$$

```{r}
x <- 1:200/20
A <- rnorm(100, 2, 0.1)
B <- rnorm(100, 1, 0.1)
error <- rnorm(100, 0, 2)
Y <- A + B * x + error

ggplot() +
  geom_point(aes(x, Y))
```

## Metropolis hastings


## Metropolis Hastings MCMC

A illustrative Bayesian approach: fitting a linear $Y=a + bx$ model with MH-MCMC

Based on https://theoreticalecology.wordpress.com/2010/09/17/metropolis-hastings-mcmc-in-r/


## Metropolis Hastings Algorithm

```{r}
trueA <- 2
trueB <- 1.5
trueSD <- 5
sample_size <- 41

## predictor values
x <- -20:20

## prediction y
y <- trueA + trueB * x + rnorm(n = length(x), mean = 0, sd = trueSD)

plot(x, y)
```



```{r}

# param = [a, b, sd]
likelihood <- function(param){
  a <- param[1]
  b <- param[2]
  sd <- param[3]

  pred <- a + b*x
  lls <- dnorm(y, mean = pred, sd = sd, log = TRUE)
  sumlls <- sum(lls)
  return(sumlls)
}
# sum(dnorm(a+bx-y, 0, sd = sd, log=TRUE))

slopevalues <- function(x){
  return(likelihood(c(trueA, x, trueSD)))
}

slopelikelihoods <- lapply(seq(from = 0, to = 7, by = 0.05), slopevalues)
plot(seq(from = 0, to = 7, by = 0.05), slopelikelihoods, type = 'l', xlab = 'values of slope parameter B')

abline(v = 1.5, col = 'red')
```

### specifying your priors

```{r}

prior <- function(param){
  a <- param[1]
  b <- param[2]
  sd <- param[3]
  aprior <- dnorm(a, mean = 0, sd = 1000, log = TRUE)
  bprior <- dgamma(b, shape = 1, rate = 0.25, log = TRUE)
#  tau     <- dgamma(1/sd, shape = 0.0001, rate = 0.00001)
#  sdprior <- 1/tau
  sdprior <- dunif(sd, min = 0, max = 1000, log = TRUE)
}

qgamma(c(0.025, 0.5, 0.975), shape = 1, rate = 0.25)
qnorm(c(0.025, 0.5, 0.975), 0, 1000)
qgamma(c(0.025, 0.5, 0.975), shape = 0.0001, rate = 0.00001)

```


```{r}
posterior <- function(param){
   post <- likelihood(param) + prior(param)
}
```

## Metropolis Hastings MCMC Algorithm

1. start somewhere reasonable
  * compute L
2. propose a new value for your params
  * compute L
3. jump to new params w/ P proportional to L(new/old)
4. if you jump, keep the old params in your basket



```{r}
proposalfunction <- function(param){
  prop <- rnorm(3, mean = param, sd = c(0.1, 0.5, 0.3))
}

mhmcmc <- function(startvalue, iterations){
  chain <- array(dim = c(iterations + 1, 3))
  chain[1,] <- startvalue
  for(i in 1:iterations){
    proposal <- proposalfunction(chain[i, ])
    probab <- exp(posterior(proposal) - posterior(chain[i, ]))
    if(runif(1) < probab){
      chain[i+1,] <- proposal
    } else {
      chain[i + 1, ] <- chain[i, ]
    }
  }
  return(chain)
}

startvalue <- c(1, 2, 3)
chain <- mhmcmc(startvalue, 10000)

mean(duplicated(chain))
burnin <- nrow(chain)/2

chain2 <- chain[burnin:nrow(chain), ]
(1:nrow(chain2)/2)*2


```


## Simulation Models

### Overview

<iframe src="https://docs.google.com/presentation/d/1uiH0PRBUw695MGV12nSpj31A7kdmz1oBwtRr0TPilQY/embed?start=false&loop=false&delayms=3000" frameborder="0" width="480" height="299" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true"></iframe>

<iframe src="https://docs.google.com/presentation/d/1OQyty2qdexakK-5zixpFH0WNAGptClyV8GQWXw7fIMA/embed?start=false&loop=false&delayms=3000" frameborder="0" width="480" height="299" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true"></iframe>

### A Photosynthesis model

Based on the coupled C4 photosynthesis - conductance model developed by Collatz and Ball Berry

> G. Collatz, M. Ribas-Carbo, J. Berry. (1992). Coupled photosynthesis-stomatal conductance model for leaves of C4 plants. Australian Journal of Plant Physiology 519–538.

Basically conductance is coupled to photosynthesis, since plants need to regulate water loss as they uptake $CO_2$:

$$g_s = m\frac{A_n h_s}{c_a}p + b$$

where $g_s$ is stomatal conducatnce, $A_n$ is net photosynthesis, $h_s$ is relative humidity, $c_a$ is $CO_2$ at leaf surface.

and Photosynthesis is

$$A_n=min(A_c, A_L)-R_d$$
Where Rubisco-limited rate is $A_c$ and RuBP limited rate is $A_L$

$$A_c=V_m\left[\frac{c_i-\Gamma}{c_i+K_c(1+O_2/K_o)}\right]$$

$$A_L=?$$
This is a non-linear equation with key plant physiological traits:

|Parameter | Description |
|:----|:-----|
| vmax      | maximum carboxylation of Rubisco according to the Collatz model.                                                 |
| alpha     | alpha parameter according to the Collatz model. Initial slope of the response to Irradiance.                     |
| kparm     | k parameter according to the Collatz model. Initial slope of the response to CO2.                                |
| theta     | theta parameter according to the Collatz model. Curvature for light response.                                    |
| beta      | beta parameter according to the Collatz model. Curvature for response to CO2.                                    |
| Rd        | Rd parameter according to the Collatz model. Dark respiration.                                                   |
| b0        | intercept for the Ball-Berry stomatal conductance model.                                                         |
| b1        | slope for the Ball-Berry stomatal conductance model.                                                             |

The rate of photosynthesis is determined by environmental factors:

|Parameter | Description |
|:----|:-----|
| Tl        | temperature of the leaf (Celsius).                                                                               |
| RH        | relative humidity (as a fraction, i.e. 0-1).                                                                     |
| Qp        | quantum flux (direct light), (micro mol m-2 s-1).                                                                |
| Catm      | Atmospheric CO2 in ppm (or micromol/mol).                                                                        |

In the end:

$$[Gs, A, C_i]=f(T, RH, Light, CO_2, v_{max}, \alpha, k, \theta, R_d, b_0, b_1)$$

### Let's run this model!

First, retrieve some meteorological data:

```{r}

metfile <- "/data/terraref/sites/ua-mac/raw_data/EnvironmentLogger/2017-05-31/2017-05-31_12-19-38_environmentlogger.json"
met <- jsonlite::fromJSON(metfile)

time <- lubridate::ymd_hms(met$environment_sensor_readings$timestamp)
par <- as.numeric(met$environment_sensor_readings$`sensor par`$value)
rh <- as.numeric(met$environment_sensor_readings$weather_station$relHumidity$value) / 100
temp <- as.numeric(met$environment_sensor_readings$weather_station$temperature$value)

```

```{r}
#devtools::install_github('ebimodeling/biocro')
library(BioCro)
```

```{r}
A <- c4photo(Qp = par, Tl = temp, RH = rh)$Assim

pairs(data.frame(A, par, temp, rh))
library(ggplot2)
ggplot()+
  geom_line(aes(time, A))
ggplot()+
  geom_line(aes(time, rh))
```

> question: is f(mean(X)) = mean(f(X))?


```{r}
testQp <- 11:20*100
testTl <- 21:30
testRH <- 21:30/50
A1 <- c4photo(Qp = mean(testQp),
              Tl = mean(testTl),
              RH = mean(testRH))
A2 <- lapply(c4photo(Qp = testQp, Tl = testTl, RH = testRH), mean)

```

Why are these different?


For non-linear functions see  Jensen's Inequality ([Wikipedia](https://en.wikipedia.org/wiki/Jensen%27s_inequality))

This means - be careful when and how you use averages - everywhere!!! Spatial and temporal downscaling is how crop modelers deal with lower resolution atmospheric model forecasts. For example the most recent IPCC 100 y climate forecasts were generated on ~100km grids ([Taylor et al 2012](http://journals.ametsoc.org/doi/pdf/10.1175/BAMS-D-11-00094.1)), thus one data point may simultaneously represent a month that is $60\degreee$F and foggy in San Fransicsco and $100\degree$F and dry in Davis, CA. At the same time, crop models need to run on local hourly data while also capturing the uncertainty represented by within and across model variability.


#### Model sensitivity

```{r}
meanQp <- mean(par)
meanTl <- mean(temp)
meanRH <- mean(rh)
plot(1:100/100, c4photo(Qp = rep(meanQp, 100),
                        Tl = rep(meanTl, 100),
                        RH = 1:100/100)$Assim,
     type = 'l', ylab = 'Assim', xlab = 'RH')

plot(1:100/4, c4photo(Qp = rep(meanQp, 100),
                      Tl = 1:100/4,
                      RH = rep(meanRH, 100))$Assim,
     type = 'l', ylab = 'Assim', xlab = 'RH')

```

#### OK. What about the other parameters?

Lets set some priors on these:

```{r}
set.seed(100)

n <- 1000
vmax <- rnorm(n, 45, 2)
Rd   <- rnorm(n, 1, 0.10)
b1   <- rnorm(n, 4, 1)

x <- 1:100
hist(vmax, probability = TRUE)
lines(x, dnorm(x, 45, 5), type = 'l')

x <- 1:200/100
hist(Rd, probability = TRUE)
lines(x, dnorm(x, 1, 0.10), type = 'l')

### sample given time series of met
A <- matrix(nrow = length(time), ncol = 1000)
for(i in 1:1000){
  A[,i] <- c4photo(Qp = par, Tl = temp, RH = rh, vmax = vmax[i], Rd = Rd[i], b1=b1[i])$Assim
}

median <- which.min(abs(quantile(colMeans(A), 0.75)-colMeans(A)))
ucl    <- which.min(abs(quantile(colMeans(A), 0.95)-colMeans(A)))
lcl    <- which.min(abs(quantile(colMeans(A), 0.25)-colMeans(A)))

ggplot() +
#  geom_smooth(aes(time, A))+
  geom_line(aes(time, A[,median])) +
  geom_line(aes(time, y = A[,lcl]), linetype = 2) +
  geom_line(aes(time, y = A[,ucl]), linetype = 2)

```
### Use met as a variable, sample over variation within the hour

```{r}

### sample over met variability

A2 <- Gs <- Ci <- Qp <- Tl <- RH <- vector(length = 1000)
for(i in 1:1000){
  j <- sample(1:length(time), size = 1)
  Qp[i] <- par[j]
  Tl[i] <- temp[j]
  RH[i] <- rh[j]
  res <- c4photo(Qp = Qp, Tl = Tl, RH = RH, vmax = vmax[i], Rd = Rd[i], b1=b1[i])
  A2[i] <- res$Assim
  Gs[i] <- res$Gs
  Ci[i] <- res$Ci
}

hist(A2)
```

Equivalent of sensitivity analysis: (where A2, Gs, Ci are response variables)

```{r}
pairs(data.frame(A2, Gs, Ci, vmax, Rd, b1, Qp, Tl, RH), pch = '.')
```

The lm is pretty much a sensitivity analysis: what is the slope of the effect of inputs on the output of the model.


```{r}

lm(A2 ~ vmax + Rd + b1 + Qp + Tl + RH)
```

The analysis of variance partitions the variance - how much if the total variance in A2 is contributed by each of the following parameters (recall that the domain for met variables is << the domain for physiological parameters ...

> What would happen if we used a whole year of meteorological data instead of the one hour of met data that we used?


```{r}
aov(A2 ~ vmax + Rd + b1 + Qp + Tl + RH)
```

```{r}
aov(Gs ~ vmax + Rd + b1 + Qp + Tl + RH)

aov(Ci ~ vmax + Rd + b1 + Qp + Tl + RH)
```

This is how you might evaluate the model for a year (2005) of climate data from Champaign, Il.

```{r illinois-met-c4photo}
data(weather05)
resA <- c4photo(Qp = weather05$solarR, Tl = weather05$DailyTemp.C -2, RH = weather05$RH)$Assim

pairs(data.frame(resA, Qp = weather05$solarR, Tl = weather05$DailyTemp.C -2, RH = weather05$RH))
```

In this case the parameter values within a plausible range are overwhelmed by the met (photosynthesis requires light and enzymes basically stop by the time they reach 0$\degree$C). This illustrates what we know about how important environment is to organisms, including crops.

### References

Taylor, K.E., R.J. Stouffer, G.A. Meehl: An Overview of CMIP5 and the experiment design.” Bull. Amer. Meteor. Soc., 93, 485-498, doi:10.1175/BAMS-D-11-00094.1, 2012.
http://journals.ametsoc.org/doi/pdf/10.1175/BAMS-D-11-00094.1

Wang et al

Miguez et al
