---
title: "7. Exploratory Data Analysis"
author: "David LeBauer"
date: '2017-05-31'
slug: exploratory-data-analysis
categories: []
tags: []
---

> **Note**: Some of the commands below, particularly those accessing the terraref database from within RStudio, require being logged into the [NDS Analytics Workbench](http://www.terraref.ndslabs.org/#/). This requires signing up for the [TERRA REF Alpha User program](https://docs.google.com/forms/d/e/1FAIpQLScBsD042RrRok70BCGCRwARTcm9etvVHqvQaz1c5X7c5y0H3w/viewform?c=0&w=1).

## Objectives

* Learning how to access data on terraref database via PostgreSQL

* Learning how to manipulate dat using the `dplyr` library, in particular:
   * [filtering, subsetting, summarizing, new variables with dplyr](http://swcarpentry.github.io/r-novice-gapminder/13-dplyr/)
   * [converting data from wide to long with tidyr](http://swcarpentry.github.io/r-novice-gapminder/14-tidyr/)

## Accessing TERRA REF via PostgreSQL

We want to access the TERRA REF database. We will do it using the [dplyr](https://cran.rstudio.com/web/packages/dplyr/README.html), library which makes it easier to connect to databases and work with data inside the database. This can be particularly useful with large datasets, where processing can be done before reading the data into memory.


```{r}
library(dplyr)
```

You can then connect to the database using (**note:** this is one of those commands that require you to be working in the NDS Analytics Workbench):

```{r}
bety_src <- src_postgres(dbname = "bety", 
                         password = 'DelchevskoOro', 
                         host = 'bety6.ncsa.illinois.edu', 
                         user = 'viewer', 
                         port = 5432)
```



#### filtering / subsetting


Let's get some data:

```{r}
traits <- tbl(bety_src, 'traits', n=Inf)

variables <- tbl(bety_src, 'variables') %>%
  mutate(variable_id = id, trait = name)

sites <- tbl(bety_src, 'sites') %>%
  mutate(site_id = id) %>% 
  dplyr::select(-geometry)

cultivars <- tbl(bety_src, 'cultivars') %>%
  mutate(cultivar_id = id, cultivar = name)
```

In the above code the symbol `%>%` is a pipe from the `dplyr` library, it passes the left hand side to the first argument slot of the right hand side.

Using `filter` we can pick what we see based on their values:

```{r}
traits_variables <- traits %>%
  left_join(variables, by = 'variable_id') %>%
  left_join(sites, by = 'site_id') %>%
  left_join(cultivars, by = 'cultivar_id') %>%
  dplyr::select(date, sitename, cultivar, trait, mean) %>% 
  filter(trait == 'canopy_height') %>%
  collect()

traits_variables_season_2 <- traits_variables %>%
      filter(grepl("Season 2",sitename))
```

### grouping and summarizing

One can combine the `group_by` and `summarize` functions to get a summary of the data we pulled:

```{r}
traits_variables_season_2 %>%
  group_by(date) %>%
  summarize(mean = mean(mean), n =n())

canopy_heights <- traits_variables_season_2 %>%
  group_by(date, cultivar) %>%
  summarize(mean = mean(mean), n =n())
```


## Plotting with ggplot

The `ggplot2` library (for more see [here](https://cran.r-project.org/web/packages/ggplot2/index.html)) can be used to create graphics.

```{r}
library(ggplot2)
```

An example using the data we pulled before:

```{r}
ggplot(data = traits_variables_season_2,
       aes(x = date, y = mean, group = cultivar)) +
  geom_smooth(se = FALSE, alpha = 0.25, size = 0.5) +
  geom_point(alpha = 0.25, size = 0.5, position = 'jitter')
```

Another example:

```{r}
### filter a subset of 25 genotypes
set.seed(1000) # to reproducibly return the same 25
height_season_2_subset <- traits_variables_season_2 %>%
  filter(cultivar %in% sample(unique(traits_variables_season_2$cultivar, 25)))

# how many unique values of cultivar_id?

#height_season_2_subset %>% group_by(cultivar_id, date) %>% summarize(n())


ggplot(data = height_season_2_subset,
       aes(x = date, y = mean, color = as.factor(cultivar))) +
  geom_smooth(se = FALSE, alpha = 0.25, size = 0.5) +
  geom_point(alpha = 0.25, size = 0.5)
```

### Faceting

Using `ggplot2` we can partition a plot into a matrix of panels. Here's an example:

```{r}
ggplot(data = height_season_2_subset,
       aes(x = date, y = mean)) +
  geom_smooth(se = FALSE) +
  geom_point() +
  facet_wrap(~cultivar) +
  ggthemes::theme_tufte()
```

### Other plots

Here is an example of a histogram:

```{r}
ggplot(data = height_season_2_subset,
       aes(mean, fill = date)) +
  geom_histogram() +
  facet_wrap(~cultivar) +
  theme_bw()
```

> Try making plots using the following functions:
> * `geom_boxplot`
> * `geom_violin`




### Visualization in R using ggplot

Some references for using ggplot:

* R for data science  [data-visualization](http://r4ds.had.co.nz/data-visualisation.html)
* [Software Carpentry 8](http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/)
